---
title: "A framework for plotting spatial objects using ggplot2"
author: "Dewey Dunnington"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggspatial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(ggspatial)
```

On first use, a GIS user who stumbles across [ggplot2](https://cran.r-project.org/package=ggplot2) will recognize much of the syntax of how plots are built from GIS language: there are layers, geometries, coordinate systems, and the ability to map attributes to the appearance of the layer (aesthetics). Using ggplot2 to plot spatial line and polygon objects has long been possible thanks to the `fortify()` implementation for `SpatialLines` and `SpatialPolygons` provided in the `ggplot2` package, and there is even an [entire wiki article](https://github.com/tidyverse/ggplot2/wiki/plotting-polygon-shapefiles) on ploting spatial polygons using ggplot2. As is, the syntax for plotting spatial objects ussing `ggplot()` is verbose and unlikely to inspire its use for the rapid creation of maps. The `ggspatial` package is designed to bridge the syntax gap between spatial objects and ggplot2 layers, allowing spatial data to be plotted efficiently.

That said, it should be noted that `ggplot2` is **not** a GIS. The grammar of graphics (outlined by Leland Wilkinson in his [excellent book](http://www.springer.com/gp/book/9780387245447)) was never designed to produce maps, and is almost infinitely flexible at (it could be argued) the expense of performance. So to qualify the above statement, it is more accurate to say that `ggspatial` is designed to leverage the existing facilities of (the `ggplot2` implementation of) the grammar of graphics to produce effecive maps with a small amount of readable code. For large data files, it is unlikely that this package will produce these plots quickly, in the computing sense. There are many packages in the [spatial task view](https://CRAN.R-project.org/view=Spatial) that are well suited to this task.

This vignette assumes that readers are familar with the usage of `ggplot2`. There are many excellent resources for learning to use `ggplot2`, one of which is the [data visualization chapter](http://r4ds.had.co.nz/data-visualisation.html) in Hadley Wickham's excellent book, [R for Data Science](http://r4ds.had.co.nz).

## Spatial Objects

There are many objects in R that represent geometries on the earth's surface (this package's definition of a "spatial object"), the most popular of which are the `Spatial*` series of objects from the [sp](https://cran.r-project.org/package=sp) package. When using `maptools::readShapeSpatial()` or `rgdal::readOGR()`, you will get a `Spatial*` object that is a `SpatialPointsDataFrame`, a `SpatialLinesDataFrame`, or a `SpatialPolygonsDataFrame`. These objects contain both the geometries (point, line, or polygon) and the attributes associated with these data. From a plotting perspective, `ggplot2` needs the geometry to know where things should plot, and it needs the attributes to map aesthetics. 

The `fortify()` method provided by `ggplot2` extracts the geometries from (some) `Spatial*` objects as a `data.frame`, but does not provide the attribues. This isn't a design flaw; in most cases attributes aren't needed and for large objects, providing them is an expensive operation that would be unnecessary in most cases. The `ggspatial` package attempts to maintain this distinction, providing previously missing `fortify()` methods for other spatial objects, and a new method (`spatial_fortify()`) that adds attribute information to the output of `fortify()` so that it can be mapped to aesthetics using `aes()`. A further method (`geom_spatial()`) is provided to produce a complete ggplot `layer()` from spatial objects, and finally, a method is provided to produce a complete `ggplot` object from a spatial object. These methods are progressively more simple to use at the expense of flexibility, since the high-level `ggpatial()` makes a number of assumptions that `fortify()` does not.

## Using fortify()

The `fortify()` method produces a `data.frame` containing the information necessary for a geometry to plot a spatial object (as a note, `fortify()` according to its documentation was intended to be used with models, but here its use applies to spatial objects). In the case of point objects, this is just a `data.frame` with columns `long`, `lat`, and `id`. Calling `fortify()` on a spatial object *always* results in the columns `long` and `lat` (even if the values may be in another coordinate system). This is partially because the existing `ggplot2` methods already did this, but is also a reminder that the values contained in these columns are not just `x` and `y` coordinates, but positions on the earth's surface. Projections are handled in `geom_spatial()`, among other functions in this package.

```{r}
# a SpatialPointsDataFrame
head(fortify(longlake_buildingsdf))
```

```{r}
# a SpatialLinesDataFrame
head(fortify(longlake_roadsdf))
```

```{r, message=FALSE}
# a SpatialPolygonsDataFrame
head(fortify(longlake_waterdf))
```

These `data.frame` outputs can then be passed to `ggplot()` with the appropriate `geom_*` and `aes()` to create a map.

```{r, message = FALSE}
ggplot() +
  geom_polypath(aes(long, lat, group = id), fortify(longlake_waterdf),
               fill = "lightblue") +
  geom_path(aes(long, lat, group = id), fortify(longlake_roadsdf)) +
  geom_point(aes(long, lat), fortify(longlake_buildingsdf),
             color = "brown", shape = 17) +
  coord_fixed()
```

(Note the use of `geom_polypath()` instead of `geom_polygon()` and the use of `geom_path()` instead of `geom_line()`. If you're curious, I suggest reading the entries for `?geom_polypath` and `?geom_path`.)

It is actually not necessary to wrap objects in `fortify()` when passing them to `ggplot()` or `geom_*` or `stat_*` functions; `ggplot2` will do this for you if you should so choose. More likely, if you need the raw data produced by `fortify()`, you will want to calculate this prior to creating the plot anyway. You will notice that the code is slightly repetitive (the mapping is always `aes(long, lat, ...)`) and doesn't allow mapping of attributes that the spatial object may contain.

This package (currently) contains `fortify()` implementations for `SpatialPoints`, `SpatialPointsDataFrame`, `SpatialLines`, and `Raster` objects. The ggplot2 package contains implementations for `Polygon`, `Polygons`, `SpatialPolygons`, `SpatialPolygonsDataFrame`, `Line`, `Lines`, and `SpatialLinesDataFrame`.

## Using spatial_fortify()

As mentioned above, `fortify()` does not include attribute information in its result. The `spatial_fortify()` function will return a `data.frame` that contains both geometry and attribute information (if applicable). For most spatial objects, there will be no attribute information. However, for `SpatialPointsDataFrame`, `SpatialLinesDataFrame`, and `SpatialPolygonsDataFrame`, there is a good chance you want to map a column that is part of your attribute table, not just your geometry. The `spatial_fortify()` function will always return a `data.frame` with columns `.long`, `.lat`, other columns returned by `fortify()` renamed to be preceeded with a `.`, and any attribute information. Non-attribute columns are preceeded by a `.` to avoid column name collisions (the column names `long`, `lat`, `piece`, `group`, and `id` returned by some `fortify()` methods could easily be attribute table names).

```{r}
head(spatial_fortify(longlake_depthdf))
```

The output of `spatial_fortify()` can then be used as input to `ggplot()` or `geom_*()` with aesthetics mapped to columns.

```{r}
ggplot() +
  geom_point(aes(.long, .lat, col = DEPTH), spatial_fortify(longlake_depthdf)) +
  coord_fixed()
```

## Using geom_spatial()

In most cases, the `geom` and `aes()` can both be guessed based on class of the object. The `geom_spatial()` function uses `spatial_fortify()`, `spatial_geom()`, and `spatial_default_aes()` to calculate the `data`, `geom`, and default `aes()` parameters that are required to create a `layer()` in ggplot2. This results in considerably cleaner code.

```{r}
ggplot() +
  geom_spatial(longlake_depthdf, aes(col = DEPTH)) +
  coord_map()
```

By default, `geom_spatial()` projects (or un-projects) your data based on the projection arguments `crsfrom` and `crsto`. **By default, data are converted to lat/lon coordinates** (the existing projection of the data is obtained from the output of `as.CRS()` or assumed to be lat/lon with a message). This is to facilitate use with `coord_map()` and `coord_quickmap()`, which already exists in the ggplot2 package. If other behaviour is desired, you can pass `NULL` (disables default lat/lon conversion) or any CRS to the `crsto` argument (can also be an EPSG code or any object from which a CRS can be extracted using `as.CRS()`). 

```{r}
ggplot() +
  geom_spatial(longlake_depthdf, aes(col = DEPTH))
```

## Using ggspatial()

So far all of the above plots have a few common elements: they initiate the plot by calling `ggplot()`, add at least one spatial layer, and then call `coord_map()` (which projects lat/lon coordinates) or `coord_fixed()` (which fixes the aspect ratio to 1). The shortcut is the `ggspatial(data, ...)` function, which is short for `ggplot() + geom_spatial(data, ...) + coord_map()`, which is the most common case for vector spatial objects.

```{r}
ggspatial(longlake_depthdf, aes(col = DEPTH))
```

## Raster objects

Raster spatial objects (of class `Raster` from the `raster` package) are treated in much the same way as vector objects, except for one major difference: spatial `Raster` objects do not have their coordinates projected by default, because `geom_raster()` throws an error if the coordinates for the raster are modified. Because of this, the shortcut method for raster spatial objects is `ggraster(data, ...)`, which is short for `ggplot() + geom_spatial(data, ...) + coord_fixed()`. The usage of `fortify()`, `spatial_fortify()`, and `geom_spatial()` are much the same.

```{r}
head(fortify(longlake_osm))
```

```{r}
head(spatial_fortify(longlake_osm))
```

```{r}
ggplot() +
  geom_spatial(longlake_osm) +
  coord_fixed()
```

```{r}
ggraster(longlake_osm)
```

The other potentially useful functions for RGB raster spatial objects (`longlake_osm` happens to be one, or you could read one using `raster::brick("...")`) are `geom_spraster_rgb()` and `annotation_spraster()`. Like `annotation_raster()` in ggplot2, `annotation_spraster()` will not adjust the extents of the plot; `geom_spraster_rgb()` will display the entire raster (in the coordinate system of the raster, whatever that may be). Using `interpolate = TRUE` will make the image appear less pixelated than the default.

```{r}
ggplot() +
  geom_spraster_rgb(longlake_osm, interpolate = TRUE) +
  coord_fixed()
```

## Open Street Map Basemaps

Using the [rosm package](https://cran.r-project.org/package=rosm), `ggspatial` can load Open Street Map tiles (among other tile sources) automatically as a backdrop for your other data (provided your data are in lat/lon, or are converted to such using `geom_spatial`'s default behaviour).

```{r}
ggplot() + 
  geom_osm() +
  geom_spatial(longlake_depthdf, aes(col = DEPTH)) +
  coord_map()
```

When using `geom_osm()`, you will *always* want to use `coord_map()` or `coord_quickmap()`. Similar to `geom_spatial()` for vector and raster objects, there is a shorthand `ggosm()` function that sets the coordinate system by default.

```{r}
ggosm() +
  geom_spatial(longlake_depthdf, aes(col = DEPTH))
```

A `geom_osm()` layer doesn't need any arguments (it uses the plot bounds to find the correct tiles to load), but a multitude of arguments can be set that control the type, zoom level, and extent that are plotted. The complete details of the arguments can be found in `?rosm::osm.image`, but perhaps the most useful argument is the first argument, which is a bounding box (or an object from which one can be extracted, see `?rosm::extract_bbox`).

```{r, message=FALSE}
ggosm("nova scotia", type = "stamenwatercolor")
```

## Extending ggspatial

There is a good chance I have missed some important use cases within this package. Spatial objects with class `ClassName` can be used with default methods provided they have a `fortify.ClassName` function defined (an S3 method for class `ClassName`). In addition, further S3 methods can be defined:

* `spatial_geom()`: Returns `GeomPoint` (point), `GeomPath` (line), `GeomPolypath` (polygon), or a custom geometry depending on the geometry that should be applied to the output of `fortify.ClassName`.
* `spatial_fortify()`: Returns a `data.frame` with columns `.long`, `.lat`, and other columns that represent attributes for each feature.
* `spatial_default_aes()`: Returns an object created with `aes()` or `aes_string()` with mappings (most likely) for `x` and `y` for the output of `spatial_fortify()`.

For advanced functionality, you can define further S3 methods:

* `spatial_stat()`: Returns the default statistic to apply (should mostly likely stay as `stat = "identity"`).
* `geom_spatial()`: Returns a ggplot2 `layer()`. Provided the above methods are defined correctly, a custom version of `geom_spatial()` should not be required.

Feel free to submit extensions to the [issues](https://github.com/paleolimbot/ggspatial/issues) page or submit a pull request.

