---
title: "A framework for plotting spatial objects using ggplot2"
author: "Dewey Dunnington"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggspatial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(ggspatial)
```

On first use, a GIS user who stumbles across [ggplot2](https://cran.r-project.org/package=ggplot2) will recognize much of the syntax of how plots are built from GIS language: there are layers, geometries, coordinate systems, and the ability to map attributes to the appearance of the layer (aesthetics). Using ggplot2 to plot spatial line and polygon objects has long been possible thanks to the `fortify()` implementation for `SpatialLines` and `SpatialPolygons` provided in the `ggplot2` package, and there is even an [entire wiki article](https://github.com/tidyverse/ggplot2/wiki/plotting-polygon-shapefiles) on ploting spatial polygons using ggplot2. As is, the syntax for plotting spatial objects ussing `ggplot()` is verbose and unlikely to inspire its use for the rapid creation of maps. The `ggspatial` package is designed to bridge the syntax gap between spatial objects and ggplot2 layers, allowing spatial data to be plotted efficiently.

That said, it should be noted that `ggplot2` is **not** a GIS. The grammar of graphics (outlined by Leland Wilkinson in his [excellent book](http://www.springer.com/gp/book/9780387245447)) was never designed to produce maps, and is almost infinitely flexible at (it could be argued) the expense of performance. So to qualify the above statement, it is more accurate to say that `ggspatial` is designed to leverage the existing facilities of (the `ggplot2` implementation of) the grammar of graphics to produce effecive maps with a small amount of readable code. For large data files, it is unlikely that this package will produce these plots quickly, in the computing sense. There are many packages in the [spatial task view](https://CRAN.R-project.org/view=Spatial) that are well suited to this task.

This vignette assumes that readers are familar with the usage of `ggplot2`. There are many excellent resources for learning to use `ggplot2`, one of which is the [data visualization chapter](http://r4ds.had.co.nz/data-visualisation.html) in Hadley Wickham's excellent book, [R for Data Science](http://r4ds.had.co.nz).

## Sample data

```{r}
load_longlake_data()
```


